---
title: "A badger bovine-TB model in R"
subtitle: "2384 badgers, 161 quarters, six imperfect tests — specified without writing Julia"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    df-print: kable
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
knitr:
  opts_chunk:
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
library(EpidemicTrajectoriesR)
```

## What this document is

A worked epidemic model, built twice.

The model is real: 2384 badgers followed over 161 quarters, four states
(`S → E → I`, with death reachable from any of them), six imperfect diagnostic
tests, capture-recapture observation, social groups that badgers move between,
and age-dependent mortality. It is the variant that reproduced the reference
implementations' mean infectious period.

Everything below is R. The rate and observation functions are ordinary R
functions; the package translates them to Julia and runs them there, so the
latent trajectory is resampled once per Gibbs sweep rather than inside every
gradient evaluation. You can see the Julia at any point with
`et_julia_source(mod)`.

Where a section shows tabs, the tabs are **alternatives** — different ways to say
the same thing, or the same model at different levels of tuning. Pick one.

::: {.callout-note collapse="true"}
## Getting set up

```{r}
#| eval: false
# install.packages("remotes")
remotes::install_github("EvoArt/EpidemicTrajectoriesR")

library(EpidemicTrajectoriesR)
et_setup()      # starts Julia and resolves its dependencies on first use
```

`et_setup()` takes a few minutes the first time and is immediate thereafter.
:::

## The data {#sec-data}

None of this is about the package — it reads the reference's CSVs and converts
two conventions. Expand it if you want to see the reshaping; skip it otherwise.

::: {.callout-note collapse="true"}
## Loading and reshaping the CSVs

Two conventions differ from the reference and are converted here:

* the reference is individual-major (`X[i, t]`); this package is time-major
  (`X[t, i]`)
* the reference encodes states sparsely (S=0, E=3, I=1, D=9); here a state is
  its position in the state space, so S=1, E=2, I=3, D=4

The one part that is not mere reshaping is the **repeat-test multiplicities**.
`TestMat.csv` carries multiple rows for the same (individual, quarter) at 1201
of 12430 cells, and 150 of those hold *contradictory* readings — genuine repeat
captures within a quarter, not an export artifact. Flattening them into one slot
(last row wins) drops real observations. Both reference implementations score
every row, and `n_neg`/`n_pos` below restore that. It matters: doing it wrong
moves the estimated mean infectious period from ~14.98 to 16.12.

```{r}
#| label: load-data
#| code-fold: true
BADGER_DATA_DIR <- Sys.getenv("BADGER_DATA_DIR", unset = file.path(
  path.expand("~"), ".julia", "dev", "EpidemicTrajectories", "badger_ref", "RData2"))

BROCK_CHANGEPOINT_RAW <- 80   # what the raw TestMat columns encode
BROCK_CHANGEPOINT     <- 101  # what we fix it to

source("../inst/examples/showcase/badger_data.R")

cat(sprintf("%d individuals x %d quarters, %d groups, %d tests\n",
            raw$n_individuals, raw$n_timepoints, raw$n_groups, raw$n_tests))
```
:::

```{r}
#| label: states
states <- c("S", "E", "I", "D")
```

The order fixes the encoding: state `k` in the trajectory means `states[k]`.

## The rates {#sec-rates}

Ordinary R functions of `(model, data, i, t)`. `model` holds the parameters;
`data` holds everything else, including the tracked arrays declared below and
anything you passed as an extra.

These come first because everything after them refers to them.

```{r}
#| label: rates
# S -> E. A group-level baseline plus a density-dependent term, with `q`
# interpolating between frequency- and density-dependence.
infection <- function(model, data, i, t) {
  g <- data$social_group[i, t]
  if (g == 0) return(0)
  infectious <- data$aggregates$n_infectious[g, t]
  alive      <- data$aggregates$n_alive[g, t]
  if (alive == 0) return(0)
  foi <- model$lambda * model$alpha[g] +
         model$beta * infectious / ((alive / data$K)^model$q)
  -expm1(-foi)
}

# E -> I. The Erlang(k, tau/k) CDF at one step, so `tau` is the mean latent
# period in quarters and `k` its shape.
progression <- function(model, data, i, t) {
  x <- data$k / model$tau
  s <- 1
  term <- 1
  for (j in 1:(data$k - 1)) {
    term <- term * x / j
    s <- s + term
  }
  1 - s * exp(-x)
}

# Gompertz-Makeham mortality, at the age the badger would reach:
#     P(survive | age) = exp(-c1 + (a2/b2) * (exp(b2*(age-1)) - exp(b2*age)))
survival_fn <- function(model, data, i, t) {
  tt <- t + 1
  age <- if (tt <= data$n_timepoints) data$age[i, tt]
         else data$age[i, data$n_timepoints] + (tt - data$n_timepoints)
  if (age < 0) return(1)
  y1 <- model$b2 * (age - 1)
  y2 <- model$b2 * age
  exp(-model$c1 + (model$a2 / model$b2) * (-exp(y1) * expm1(y2 - y1)))
}
```

Where an animal starts, for one already alive when watching began:

```{r}
#| label: starting-state
starting_state <- function(model, data, X, i, t) {
  p <- numeric(data$n_states)
  st <- data$sampling_period[i][1]
  nuE <- 0
  nuI <- 0
  if (data$birth_time[i] < st) {
    idx <- 0
    for (j in 1:length(data$nu_times)) {
      if (data$nu_times[j] == st) idx <- j
    }
    if (idx > 0) {
      nuE <- model$nu[idx, 1]
      nuI <- model$nu[idx, 2]
    }
  }
  p[1] <- 1 - nuE - nuI
  p[2] <- nuE
  p[3] <- nuI
  p[4] <- 0
  p
}
```

## Declaring the transitions {#sec-transitions}

This is the part with the most freedom, so it is worth seeing the alternatives
side by side. **All three tabs describe the same model.** They differ only in how
much you write out and how much the package fills in.

The first tab is the one this document carries forward; the other two are shown
for comparison and are not run.

::: {.panel-tabset}

### With `et_survival()` (used here)

Declare the two disease moves and one survival function. `et_survival()` then
does two things: it scales every non-death rate by the survival probability, and
it gives **every** live state a transition to `"D"` with the remaining mass —
including `I → D`, which is never written here.

```{r}
#| label: trans-survival
trans <- et_transitions(states,
  survival = et_survival(survival_fn, death = "D"),
  "S -> E" = infection,
  "E -> I" = progression)
```

Two moves declared, five generated: `S→E`, `E→I`, `S→D`, `E→D`, `I→D`. The
self-transitions (`S→S`, `E→E`, `I→I`) take whatever probability is left over.

Use this whenever mortality is a competing risk acting on every live state —
which is the usual case for a survival-structured model.

### Death written out by hand

The same model with no `et_survival()`. Every transition into death is declared
explicitly, and each disease rate carries its own survival factor.

```{r}
#| label: trans-manual
#| eval: false
death_rate <- function(model, data, i, t) 1 - survival_fn(model, data, i, t)

infection_x_survival <- function(model, data, i, t) {
  survival_fn(model, data, i, t) * infection(model, data, i, t)
}
progression_x_survival <- function(model, data, i, t) {
  survival_fn(model, data, i, t) * progression(model, data, i, t)
}

trans_manual <- et_transitions(states,
  "S -> E" = infection_x_survival,
  "E -> I" = progression_x_survival,
  "S -> D" = death_rate,
  "E -> D" = death_rate,
  "I -> D" = death_rate)
```

Identical results, five declarations instead of two, and `I → D` is now yours to
remember. Worth it only when death is not a uniform competing risk — say if
being infectious changed the *form* of the mortality model rather than just its
rate.

### Self-transitions written out

By default each state's self-transition takes the leftover probability mass, so
rows sum to one. You can turn that off and declare them yourself:

```{r}
#| label: trans-noself
#| eval: false
stay_S <- function(model, data, i, t) {
  survival_fn(model, data, i, t) * (1 - infection(model, data, i, t))
}
stay_E <- function(model, data, i, t) {
  survival_fn(model, data, i, t) * (1 - progression(model, data, i, t))
}
stay_I <- function(model, data, i, t) survival_fn(model, data, i, t)
stay_D <- function(model, data, i, t) 1

trans_noself <- et_transitions(states,
  auto_self = FALSE,
  "S -> E" = infection_x_survival,
  "E -> I" = progression_x_survival,
  "S -> S" = stay_S,
  "E -> E" = stay_E,
  "I -> I" = stay_I,
  "D -> D" = stay_D,
  "S -> D" = death_rate,
  "E -> D" = death_rate,
  "I -> D" = death_rate)
```

Ten declarations for the same four-state model, and now **you** are responsible
for each row summing to one — an arithmetic slip gives an improper transition
matrix rather than an error.

Reach for this only when a self-transition is genuinely not the residual
probability: an absorbing state with its own dynamics, or a model where staying
put has a rate that is not one minus everything else.

:::

## Tracked arrays {#sec-aggregates}

The force of infection needs to know how many groupmates are infectious and how
many are alive. You declare the arrays and how they update; the package
allocates them and keeps them in step with the trajectory.

The reason it needs the *update* rather than just the array is that the sampler
must be able to take one individual's contribution back out, resample it, and
put it back — which is what gives each individual the leave-one-out count its
force of infection wants. So every update has to be reversible.

The tuned build later needs two extra arrays, so both versions are shown. The
second is the one carried forward.

::: {.panel-tabset}

### Derived reverse (enough for the plain build)

Write the update in the shape `A[idx] <- A[idx] + contribution` and the package
derives the reverse for you. It cannot be got wrong.

```{r}
#| label: aggs-simple
#| eval: false
aggs_simple <- et_aggregate(
  states,
  arrays = list(
    n_infectious = et_array("Int", c(raw$n_groups, raw$n_timepoints)),
    n_alive      = et_array("Int", c(raw$n_groups, raw$n_timepoints))),
  update = function(model, data, X, state, i, t) {
    if (data$social_group[i, t] > 0) {
      n_infectious[data$social_group[i, t], t] <-
        n_infectious[data$social_group[i, t], t] + (state == "I")
    }
    if (data$social_group[i, t] > 0) {
      n_alive[data$social_group[i, t], t] <-
        n_alive[data$social_group[i, t], t] + (state != "D")
    }
  })
```

Anything outside that shape is refused rather than guessed at — a wrong reverse
would silently desynchronise the arrays from the trajectory.

### Adding a hand-written reverse (used here)

Some updates cannot be expressed that way. The tuned coupling below needs counts
of how many susceptible badgers went on to `E` (and to `S`) at each step — a
contribution that depends on where the individual goes **next**, which the
`A[i] <- A[i] + x` shape does not expose.

`et_summary()` takes the whole update, including its reverse. `et_aggregates()`
mixes the two kinds:

```{r}
#| label: aggs-hand
aggs <- et_aggregates(
  states,
  arrays = list(
    n_infectious = et_array("Int", c(raw$n_groups, raw$n_timepoints)),
    n_alive      = et_array("Int", c(raw$n_groups, raw$n_timepoints)),

    nSE = et_summary("Int", c(raw$n_groups, raw$n_timepoints),
      update = function(model, data, X, state, i, t, reverse) {
        g <- data$social_group[i, t]
        if (g > 0 && t < data$n_timepoints) {
          contrib <- (state == "S") && (X[t + 1, i] == "E")
          if (reverse) {
            nSE[g, t] <- nSE[g, t] - contrib
          } else {
            nSE[g, t] <- nSE[g, t] + contrib
          }
        }
      }),

    nSS = et_summary("Int", c(raw$n_groups, raw$n_timepoints),
      update = function(model, data, X, state, i, t, reverse) {
        g <- data$social_group[i, t]
        if (g > 0 && t < data$n_timepoints) {
          contrib <- (state == "S") && (X[t + 1, i] == "S")
          if (reverse) {
            nSS[g, t] <- nSS[g, t] - contrib
          } else {
            nSS[g, t] <- nSS[g, t] + contrib
          }
        }
      })),

  update = function(model, data, X, state, i, t) {
    if (data$social_group[i, t] > 0) {
      n_infectious[data$social_group[i, t], t] <-
        n_infectious[data$social_group[i, t], t] + (state == "I")
    }
    if (data$social_group[i, t] > 0) {
      n_alive[data$social_group[i, t], t] <-
        n_alive[data$social_group[i, t], t] + (state != "D")
    }
  })
```

You are now responsible for the reverse exactly undoing the forward update.
Nothing checks it, and getting it wrong fails silently.

::: {.callout-warning}
`state == "S"` and `X[t + 1, i] == "E"` look alike and are **not** the same
comparison. `state` is the state being applied, so it compares by name;
`X[t + 1, i]` is the trajectory, which holds integer codes, so the name resolves
to its index. The package handles both, and refuses a name your state space does
not contain — otherwise that comparison would simply never match.
:::

:::

## The observation process {#sec-observation}

Capture × tests. A badger that was not caught is not tested; one caught later
cannot have been dead now.

The weight for a state is a **product of two factors**: whether the badger was
caught, and what its tests said. That product is why the next section can split
them.

```{r}
#| label: obs-tests
# Tests only, one state at a time. No array allocated per cell.
#
# Each READING is scored once, via the repeat-capture multiplicities from
# @sec-data. theta is sensitivity in I, theta*rho in E, phi is specificity in S.
#
# et_one() rather than 1: `w` accumulates a product of test-parameter terms, so
# it must start at the parameter scalar type. Starting at an integer would make
# the accumulator a union type the moment a parameter multiplied into it.
tests_weight <- function(model, data, X, i, t, s) {
  if (data$capture[t, i] == 0) return(1)
  if (s == 4) return(1)
  w <- et_one()
  for (j in 1:dim(data$tests)[3]) {
    nneg <- data$n_neg[t, i, j]
    npos <- data$n_pos[t, i, j]
    if (nneg + npos > 0) {
      if (s == 1) {
        phi <- model$phis[j]
        if (npos > 0) w <- w * (1 - phi)^npos
        if (nneg > 0) w <- w * phi^nneg
      } else if (s == 2) {
        tr <- model$thetas[j] * model$rhos[j]
        if (npos > 0) w <- w * tr^npos
        if (nneg > 0) w <- w * (1 - tr)^nneg
      } else {
        th <- model$thetas[j]
        if (npos > 0) w <- w * th^npos
        if (nneg > 0) w <- w * (1 - th)^nneg
      }
    }
  }
  w
}

# capture x tests: the whole weight for state `s`.
observation <- function(model, data, X, i, t, s) {
  eta <- model$etas[data$season[t]]
  capture_w <- if (data$capture[t, i] == 0) {
    if (s == 4) ifelse(t <= data$last_capture_time[i], 0, 1) else 1 - eta
  } else {
    ifelse(s == 4, 0, eta)
  }
  capture_w * tests_weight(model, data, X, i, t, s)
}
```

### Why the likelihood scores only one of the two factors

The **filter** needs both factors. It must know that a non-capture is
informative, and that a badger caught later cannot have been dead earlier.

The **likelihood** must score the tests factor *only*. `etas` — the capture
probabilities — is drawn by an exact conjugate kernel, so if the likelihood also
scored capture, `etas` would be counted twice: once by its own kernel, and again
through the gradient.

Because the weights **multiply**, dropping a factor from the likelihood drops
exactly its term and nothing else. That is why the two are separate arguments,
and it is exact rather than an approximation:

```{r}
#| label: obs-split-note
#| eval: false
# in et_data():
#   observation_weight = observation      # filter:     capture x tests
#   likelihood_weight  = tests_weight     # likelihood: tests only
```

This is also what keeps `etas` out of the derived dependency list — see
@sec-depends.

## The tuned coupling {#sec-coupling}

This is the one place the R interface hands over to Julia, deliberately. The
closed form assumes a single coupled transition with a per-group
frequency-dependent force of infection — a modelling choice that cannot be
inferred from the spec.

`et_julia()` accepts literal Julia anywhere a transpiled R function is accepted,
and reads the same tracked arrays the R code declared above.

```{r}
#| label: coupling
coupling <- et_julia('
    t == data.n_timepoints && return ones(n_states)
    g = data.social_group[i, t]
    g == 0 && return ones(n_states)
    I_minus = data.aggregates.n_infectious[g, t]
    M_minus = data.aggregates.n_alive[g, t]
    nSE, nSS = data.aggregates.nSE[g, t], data.aggregates.nSS[g, t]
    logw = zeros(Float64, n_states)
    @inbounds for s in 1:n_states
        # The focal was reversed out of the aggregates for its own resample, so
        # its own contribution to the denominator is added back per candidate.
        I, M = s == 3 ? (I_minus + 1, M_minus + 1) :
               s == 4 ? (I_minus, M_minus) : (I_minus, M_minus + 1)
        foi = M == 0 ? 0.0 :
              -expm1(-(model.lambda * model.alpha[g] +
                       model.beta * I / ((M / data.K)^model.q)))
        logw[s] = nSE * log(max(foi, 1e-12)) + nSS * log(max(1.0 - foi, 1e-12))
    end
    logw .-= maximum(logw)          # stabilise before exponentiating
    return exp.(logw)
')
```

The default coupling recomputes, for each candidate state of the focal
individual, how probable every affected neighbour's realised move is. Keeping
per-(group, time) running totals of the `S → E` and `S → S` moves turns that
into a closed form in those totals, with no neighbour loop at all.

## Assembling {#sec-assemble}

Everything the package does not name is yours: pass it in `extras` and read it
back as `data$whatever` inside any function above.

The four tuning choices are marked. Each is exact, not an approximation.

```{r}
#| label: build-data
dat <- et_data(
  n_individuals  = raw$n_individuals,
  n_timepoints   = raw$n_timepoints,
  transitions    = trans,
  starting_state = starting_state,
  aggregates     = aggs,
  observation_weight  = observation,      # filter:     capture x tests
  likelihood_weight   = tests_weight,     # likelihood: tests only    (2)
  rest_contribution   = coupling,         # O(n_states) coupling      (3)
  coupled_transitions = list(c("S", "E")),                         # (1)
  sampling_period     = raw$sampling_period,
  # Badgers move between groups, so who each one affects varies with time.
  affected_individuals = et_affected_from_groups(raw$social_group),
  helpers = list(et_helper(tests_weight, "tests_weight")),
  extras = list(
    social_group = raw$social_group, age = raw$age, capture = raw$capture,
    capt_effort = raw$capt_effort, tests = raw$tests,
    n_neg = raw$n_neg, n_pos = raw$n_pos, season = raw$season,
    birth_time = raw$birth_time, nu_times = raw$nu_times,
    last_capture_time = raw$last_capture_time,
    first_capture_time = raw$first_capture_time,
    K = raw$K, k = raw$k))
```

1. **`coupled_transitions`** — the only way one badger changes another's fate is
   by contributing to its force of infection. A neighbour whose realised move is
   not `S → E` has the same probability whatever the focal does, so it cancels
   on normalisation and can be skipped.
2. **`likelihood_weight`** — one entry of the weight vector rather than the whole
   array, and here a different factor besides (@sec-observation).
3. **`rest_contribution`** — the closed form from @sec-coupling.
4. **The `depends=` split** — @sec-depends. You do not write it.

Now the parameters and priors. `et_exponential()` takes a **rate**; Julia's
`Exponential` is parameterised by scale, and the package converts. `nu` is owned
by a conjugate kernel, so it is declared `"latent"`: a placeholder density, with
the real draw coming from the kernel.

```{r}
#| label: build-model
mod <- et_model(
  data = dat,
  parameters = list(
    tau    = et_par(et_exponential(1 / 100), init = 5.0),
    alpha  = et_par(et_exponential(1), init = rep(0.5, raw$n_groups),
                    n = raw$n_groups),
    lambda = et_par(et_exponential(1), init = 0.5),
    beta   = et_par(et_exponential(1), init = 0.3),
    q      = et_par(et_beta(1, 1), init = 0.2),

    c1 = et_par(et_exponential(1), init = 0.45),   # Makeham constant
    a2 = et_par(et_exponential(1), init = 0.05),   # Gompertz scale
    b2 = et_par(et_exponential(1), init = 0.3),    # Gompertz rate

    thetas = et_par(et_beta(1, 1), init = rep(0.3, raw$n_tests), n = raw$n_tests),
    rhos   = et_par(et_beta(1, 1), init = rep(0.5, raw$n_tests), n = raw$n_tests),
    phis   = et_par(et_beta(1, 1), init = rep(0.5, raw$n_tests), n = raw$n_tests),
    etas   = et_par(et_beta(1, 1), init = rep(0.3, raw$n_seasons),
                    n = raw$n_seasons),
    nu     = et_par(init = rep(0.05, raw$n_nu_times * 2),
                    dim = c(raw$n_nu_times, 2), kind = "latent")),
  # Entry gate: a badger is known alive at first capture, but its disease
  # dynamics before then were not watched. The likelihood scores the disease
  # moves pre-entry and divides the survival factor back out.
  entry_time = "first_capture_time")
```

And the sampler. `et_adaptive_hmc()` is a fixed-trajectory HMC block that adapts
its mass matrix and step size during warm-up — exactly as NUTS does — and then
holds both fixed for sampling. Every iteration costs a constant `n_steps`
leapfrog steps, so the run has a flat, predictable cost and never doubles a tree.

`n_steps = 15` matches the C++ reference's *expected* trajectory length (it draws
L uniform on 1..30, mean 15.5), not its nominal 30. `metric = "dense"` learns
correlations between parameters, which costs more warm-up but suits a posterior
where the epidemic parameters trade off against one another.

::: {.callout-warning}
## Why not a hand-tuned metric

`et_hmc()` takes a fixed metric you supply. That has to be right in both **order**
and **scale**, and a scale far too small produces a chain that **never moves**
while every acceptance diagnostic still looks healthy — silence rather than an
error.

That is not hypothetical: the hand-tuned metric originally used here was carried
over from a different variant of this model and left the block completely frozen,
with `sd = 0` on every parameter. Adapting removes the failure mode.
:::

```{r}
#| label: build-blocks
blocks <- list(
  et_adaptive_hmc(c("tau", "alpha", "lambda", "beta", "q", "c1", "a2", "b2",
                    "thetas", "rhos", "phis"),
                  target_accept = 0.8, n_steps = 15, metric = "dense"),

  # Closed-form conditionals: an exact independent draw beats a correlated HMC
  # step, so these do not go in the block above.
  et_conjugate_capture_prob("etas",
    caught = "capture", effort = "capt_effort", group = "social_group",
    index  = "season", dead_state = "D", n = raw$n_seasons),

  et_conjugate_initial_state("nu",
    at = "nu_times", states = c("S", "E", "I"), n = raw$n_nu_times,
    eligible = function(X, data, i, t) {
      data$sampling_period[i][1] == t && data$birth_time[i] < t
    }),

  et_iffbs("X"))
```

::: {.callout-note collapse="true"}
## The plain build, for comparison

Everything above is the tuned build. The plain one differs in exactly four
lines: it hands `et_data()` a vector-returning `observation_process` instead of
the split weights, drops `rest_contribution` and `coupled_transitions`, uses
`aggs_simple` (no `nSE`/`nSS`), and swaps the hand-tuned HMC block for
`et_nuts()`.

```{r}
#| label: build-friendly
#| eval: false
dat_plain <- et_data(
  n_individuals  = raw$n_individuals,
  n_timepoints   = raw$n_timepoints,
  transitions    = trans,
  starting_state = starting_state,
  aggregates     = aggs_simple,
  observation_process  = observation_vector,   # the vector-returning form
  sampling_period      = raw$sampling_period,
  affected_individuals = et_affected_from_groups(raw$social_group),
  extras = list(
    social_group = raw$social_group, age = raw$age, capture = raw$capture,
    capt_effort = raw$capt_effort, tests = raw$tests,
    n_neg = raw$n_neg, n_pos = raw$n_pos, season = raw$season,
    birth_time = raw$birth_time, nu_times = raw$nu_times,
    last_capture_time = raw$last_capture_time,
    first_capture_time = raw$first_capture_time,
    K = raw$K, k = raw$k))

blocks_plain <- list(
  et_nuts(c("tau", "alpha", "lambda", "beta", "q", "c1", "a2", "b2",
            "thetas", "rhos", "phis")),
  et_conjugate_capture_prob("etas",
    caught = "capture", effort = "capt_effort", group = "social_group",
    index  = "season", dead_state = "D", n = raw$n_seasons),
  et_conjugate_initial_state("nu",
    at = "nu_times", states = c("S", "E", "I"), n = raw$n_nu_times),
  et_iffbs("X"))
```

You could drop `blocks` entirely: every parameter would go into one NUTS block
and the trajectory would still get iFFBS. It is spelled out because the two
conjugate kernels are worth having.

The full plain build is in `inst/examples/showcase/badger_friendly.R`.
:::

## Dependencies you do not have to declare {#sec-depends}

Each Gibbs block can skip likelihood terms its parameters cannot move. That is a
large saving and a sharp edge: getting the list wrong drops a real gradient
contribution with **no error and no warning**.

You do not write the list. The package reads which `model$x` each of your
functions touches — following your helpers — and emits it:

```{r}
#| label: depends
#| eval: false
src <- et_julia_source(mod, blocks)
src$depends
```

For the two builds above:

| build | epidemic term | observation term |
|---|---|---|
| friendly | `tau alpha lambda beta q c1 a2 b2 nu X` | `thetas rhos phis` **`etas`** `X` |
| tuned | `tau alpha lambda beta q c1 a2 b2 nu X` | `thetas rhos phis X` |

The difference is the whole point of splitting the observation process.
`etas` reaches the *filter*, which is never differentiated, and never reaches
the likelihood — so it drops out, and cannot be double-counted against its own
conjugate kernel. Nobody wrote either list.

Verify before trusting a gradient:

```{r}
#| label: check-depends
#| eval: false
dep <- et_check_depends(mod, blocks, x_init = raw$X_init)
subset(dep, verdict != "ok")
```

This perturbs each parameter and checks that no term it was excluded from
actually moves. It is a probe rather than a proof — a dependency inactive at the
tested values would pass — but it turns a silent failure into a loud one.

## Fitting {#sec-fit}

```{r}
#| label: fit
#| eval: false
et_setup()

# A cheap structural check first: a non-finite log density means the model is
# misspecified before any sampler is involved.
ll <- et_loglik(mod, blocks, x_init = raw$X_init)
stopifnot(all(is.finite(ll)))

fit <- et_sample(mod, blocks,
                 n_sweeps = 5000, n_adapts = 1000, seed = 13,
                 x_init = raw$X_init,
                 save_x = "badger_X.jld2", save_every = 100)

summary(fit)
```

::: {.callout-important}
`save_x` streams the latent trajectory to disc. It is 161 × 2384 integers
**every sweep**, so it is always kept live for conditioning but never carried in
the chain object. Without `save_x` the per-sweep copy is simply dropped, which
is the right default — you only need it for the residuals below.

**`save_every` is a flush interval, not thinning.** Every sweep is archived
either way; `save_every` only sets how many are buffered before a file is
written, and each flush gets its own file. At 5000 sweeps this model's archive is
therefore about **15 GB** (161 × 2384 `Int64` = 3.1 MB per draw), across 50
files. There is no thinning option in the stack today, so either budget the disc
or use [`et_collect()`](#sec-online), which scores the residuals during the fit
and never stores a trajectory at all.
:::

## Residuals {#sec-residuals}

A fit tells you what the parameters are. A residual tells you whether the model
that produced them is any good.

Each is a PIT residual: **Uniform(0, 1) if the mechanism it targets is right**.
A departure from uniformity therefore points at that specific mechanism, rather
than saying "the model is bad" in general.

::: {.panel-tabset}

### None

Perfectly reasonable. Nothing extra to declare, and `et_sample()` drops the
per-sweep trajectory rather than storing it.

```{r}
#| label: resid-none
#| eval: false
fit <- et_sample(mod, blocks, n_sweeps = 5000, n_adapts = 1000, seed = 13,
                 x_init = raw$X_init)
summary(fit)
```

### Post-hoc (recommended)

Archive the trajectory during the fit, score it afterwards. New residuals can be
added later without refitting.

```{r}
#| label: resid-posthoc
#| eval: false
fit <- et_sample(mod, blocks, n_sweeps = 5000, n_adapts = 1000, seed = 13,
                 x_init = raw$X_init,
                 save_x = "badger_X.jld2", save_every = 100)

residuals_spec <- list(
  # How long a badger LIVES, against Gompertz-Makeham.
  et_residual_survival(
    origin       = "birth_time",
    condition_on = "first_capture_time",
    censor_at    = "last_capture_time",
    death        = "D", name = "survival"),

  # How long a SUSCEPTIBLE badger waits before infection.
  et_residual_waiting("S", "E",
    origin = "window_start", censor_at = c("window_end", "D"),
    accumulate = "discrete_product", name = "exposure"),

  # How long an EXPOSED badger takes to become infectious - the residual for tau.
  et_residual_waiting("E", "I",
    origin = "entry_to_from_state", censor_at = c("window_end", "D"),
    accumulate = "discrete_product", name = "latent_period"))

res <- et_residuals(fit, residuals_spec, sync_aggregates = TRUE)
et_residual_summary(res)
```

`sync_aggregates` rebuilds the tracked arrays for each draw before scoring it.
It is needed here because the force of infection reads them, so a spec-derived
hazard depends on them agreeing with *that draw's* trajectory rather than
whatever was current when the fit ended.

### Streamed during sampling {#sec-online}

When the trajectories are too large to keep even on disc, score them as they are
produced and never store them. `et_collect()` fits and scores in one call:

```{r}
#| label: resid-stream
#| eval: false
out <- et_collect(mod, blocks, residuals_spec,
                  n_sweeps = 5000, n_adapts = 1000, seed = 13,
                  x_init = raw$X_init,
                  n_keep = 500)

et_residual_summary(out$residuals)
summary(out$fit)
```

Storage is `n_individuals x n_keep` doubles per residual — for this model about
9.5 MB each, against the ~15 GB of trajectories `save_x` would archive. Nothing
is written to disc and no trajectory is ever kept.

The trade is that the residuals must be decided **before** the fit. Archiving
lets you add new ones afterwards without refitting; this does not. So use
`save_x` while you are still deciding what to look at, and `et_collect()` once
you know.

:::

### Reading the result

`et_residual_summary()` reports the mean and the share of values in each
quartile:

| | what it means |
|---|---|
| mean ≈ 0.5, quartiles ≈ 0.25 each | no evidence of misfit in that mechanism |
| mean pushed high | events happening **later** than the model expects |
| mean pushed low | events happening **earlier** than the model expects |
| mass piled at one end | the mechanism is wrong in shape, not just scale |

::: {.callout-warning}
Residuals from a short or unconverged chain mean nothing. They are diagnostics
of a *fitted* model, so read them only after the chain has converged.
:::

::: {.callout-note collapse="true"}
## Are the residuals tied to this model?

No. `death = "D"` is a name resolved against your declared state space, not a
constant — a model with `c("Alive", "Dead")` passes `death = "Dead"`, and one
with no death state does not use that residual at all.

The waiting-time residual is more general still: it derives its per-step hazard
from your transitions, so **any** declared move gets a residual for free. `S → E`,
`E → I`, `I → R`, `Exposed → Symptomatic` — no new code.

One thing genuinely has no default. `accumulate` must be stated, because
`"discrete_product"` and `"cumulative_hazard"` agree only when the per-step
probability is `1 - exp(-hazard)`. Choosing wrong is a silent miscalibration, so
the package will not guess.
:::

## Who infected whom: R~i~ {#sec-ri}

A residual asks whether the model fits. This asks something else: **how many
secondary infections did each badger cause?**

```{r}
#| label: reproduction-numbers
#| eval: false
# The force of infection, split into the parts you want attributed separately.
# These are ordinary rate functions -- the same subset, the same
# (model, data, i, t) contract as anything in et_transitions().
background <- function(model, data, i, t) {
  g <- data$social_group[i, t]
  if (g == 0) return(0)
  model$lambda * model$alpha[g]
}

transmission <- function(model, data, i, t) {
  g <- data$social_group[i, t]
  if (g == 0) return(0)
  alive <- data$aggregates$n_alive[g, t]
  if (alive == 0) return(0)
  model$beta * data$aggregates$n_infectious[g, t] / ((alive / data$K)^model$q)
}

R_i <- et_reproduction_numbers(fit,
  components = list(background = background, transmission = transmission),
  secondary  = "transmission",     # the part attributable to other badgers
  infection  = c("S", "E"),
  infectious_state = "I")

aggregate(value ~ individual, R_i, mean, na.rm = TRUE) |> head()
```

::: {.callout-note}
## Why you declare the split rather than the package deriving it

`S → E` is **one** rate function returning **one** number, and nothing in it says
which part is background and which is transmission. The package cannot take it
apart, so you name the components and say which one is transmission. It sums
them, normalises, and never needs to know what any of them means — the same
bargain as aggregates and rates.

That also buys generality for free: the split is defined for any number of
components, so per-neighbour attribution ("which badger infected this one") is
just a longer list.
:::

For each infection, the transmission share of the victim's force of infection is
divided among the infectives in its group. Background infections are attributed
to nobody — that is the point of the split. A badger that was never infectious
gets `NA`: it never had the opportunity, which is different from having had it
and infected nobody (a genuine `0`).

::: {.callout-important}
Unlike every other quantity here, R~i~ is **not** computed one individual at a
time — attributing an infection needs every infective present in the group at
that moment. So it takes a whole-population pass and has its own entry point
rather than joining `et_residuals()`.
:::

## Seeing the Julia {#sec-julia}

The generated module is a first-class artefact, not a hidden intermediate. It is
commented, it runs standalone, and it is what to show a Julia-literate colleague
when something goes wrong.

```{r}
#| label: source
#| eval: false
et_julia_source(mod, blocks)   # print the module
et_source(fit)                 # the module behind a fit
et_eval("...")                 # run anything in the session
```

## Reproducing the reference {#sec-reference}

This is the **Gompertz-Makeham** variant with the **repeat-test fix** — the pair
that reproduced the reference implementations:

| implementation | mean infectious period |
|---|---|
| upstream C++ | 15.06 |
| **this model** | **14.98** (sd 1.76, ESS 403, MCSE 0.088, Rhat 1.003) |
| semi-Markov reference | 14.87 |

Both differences are inside Monte Carlo error. Two earlier variants were not:
before the repeat-test fix it read 16.12, and before the changepoint fix, 14.10.

::: {.callout-important}
## One limitation, stated plainly

The Brock changepoint `ξ` is sampled in the reference by a bespoke Metropolis
kernel — an in-place swap of the test columns plus a windowed acceptance ratio —
which the R interface cannot yet express. This document **fixes it at 101**.

The posterior for `ξ` visits only 100–102 (98.9% at 100–101), so that is close to
its mode, but it is a fixed-`ξ` run rather than the sampled-`ξ` run that produced
the number above.
:::
