---
title: "A cattle E. coli model, in R and in Julia"
subtitle: "The same model written both ways, side by side"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    df-print: kable
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
knitr:
  opts_chunk:
    comment: "#>"
---

## What this document is

One small model, written twice: once in R, once in Julia. Every code block below
has both, and **the tabs are the same model** — pick whichever language you work
in.

The point is not that the two look similar (they do), but that the *shape* of the
specification is identical. If you learn it in one language you have learned it
in the other, and a model prototyped in R can graduate to Julia without being
rethought.

The model is deliberately small and readable. Where the badger document
optimises, this one does not: it is about how little you have to write.

### The model

Two states, `S` and `I`, with recurrent `S ⇄ I`. An animal is infected at a rate
driven by how many of its penmates are infected, and recovers back to
susceptible.

Neither state is ever observed. All we see are two imperfect diagnostic tests,
on a subset of days — so the trajectory is entirely latent, and inferring it is
the whole problem.

This is the E. coli cattle model of Touloupou et al. (2019).

## Setting up

::: {.panel-tabset}

### R

```{r}
#| eval: false
# install.packages("remotes")
remotes::install_github("EvoArt/EpidemicTrajectoriesR")

library(EpidemicTrajectoriesR)
et_setup()      # starts Julia and resolves its dependencies on first use
```

### Julia

```julia
using Pkg
Pkg.add(url = "https://github.com/EvoArt/EpidemicTrajectories.jl")
Pkg.add(url = "https://github.com/EvoArt/PracticalBayes.jl")
Pkg.add(url = "https://github.com/EvoArt/PracticalEpiBayes.jl")

using EpidemicTrajectories, PracticalBayes, PracticalEpiBayes
using Distributions, Random, StableRNGs
import AbstractMCMC
```

:::

## The problem size

::: {.panel-tabset}

### R

```{r}
#| label: size-r
n_pens        <- 10
n_per_pen     <- 8
n_timepoints  <- 80
n_individuals <- n_pens * n_per_pen
group         <- rep(seq_len(n_pens), each = n_per_pen)
observed_days <- seq(1, n_timepoints, by = 6)

states <- c("S", "I")
```

### Julia

```julia
n_pens        = 10
n_per_pen     = 8
n_timepoints  = 80
n_individuals = n_pens * n_per_pen
group         = repeat(1:n_pens; inner = n_per_pen)
observed_days = collect(1:6:n_timepoints)

state_space = [:S, :I]
```

:::

Declaring the state space fixes the numbering, so you and the package agree on
which state is which: `S` is 1, `I` is 2.

## What to track

The force of infection needs to know how many penmates are infected. Declare the
array and how it updates; the package allocates it and keeps it in step with the
trajectory.

It needs the *update* rather than just the array because the sampler must be
able to take one animal's contribution back out, resample it, and put it back —
which is what gives each animal the leave-one-out count its force of infection
wants.

::: {.panel-tabset}

### R

```{r}
#| label: aggs-r
aggs <- et_aggregate(
  states,
  arrays = list(n_infected = et_array("Int", c(n_pens, n_timepoints))),
  update = function(model, data, X, state, i, t) {
    n_infected[data$group[i], t] <- n_infected[data$group[i], t] + (state == "I")
  })
```

### Julia

```julia
aggs = @aggregate state_space begin
    @array n_infected Int (n_pens, n_timepoints)
    n_infected[data.group[i], t] += (state == :I)
end
```

:::

The package attaches no meaning to this array. It is whatever your rates need.

## The rates

Ordinary functions of `(model, data, i, t)`. `model` holds the parameters; `data`
holds everything else, including the tracked array above.

::: {.panel-tabset}

### R

```{r}
#| label: rates-r
infection <- function(model, data, i, t) {
  I_minus <- data$aggregates$n_infected[data$group[i], t]
  -expm1(-(model$alpha + model$beta * I_minus))
}

recovery <- function(model, data, i, t) 1 / model$m
```

### Julia

```julia
function infection(model, data, i, t)
    I_minus = data.aggregates.n_infected[data.group[i], t]
    -expm1(-(model.α + model.β * I_minus))
end

recovery(model, data, i, t) = 1 / model.m
```

:::

`n_infected` excludes the animal currently being resampled, which is exactly the
leave-one-out count the force of infection needs — you get that for free from
having declared a reversible update.

## The transitions

::: {.panel-tabset}

### R

```{r}
#| label: trans-r
trans <- et_transitions(states,
  "S -> I" = infection,
  "I -> S" = recovery)
```

### Julia

```julia
trans_mat = @transitions state_space begin
    S -> I = infection
    I -> S = recovery
end
```

:::

Two moves declared. The self-transitions (`S→S`, `I→I`) take whatever
probability is left over, so the rows sum to one without your having to say so.

## Where an animal starts

::: {.panel-tabset}

### R

```{r}
#| label: start-r
starting_state <- function(model, data, X, i, t) {
  p <- numeric(data$n_states)
  p[1] <- 1 - model$nu
  p[2] <- model$nu
  p
}
```

### Julia

```julia
starting_state(model, data, X, i, t) = [1 - model.ν, model.ν]
```

:::

## What we observe

Two imperfect tests, each with its own sensitivity and perfect specificity. A
negative entry means the animal was not tested that day, so the result says
nothing.

The package knows none of this — it just multiplies these weights into the
filter.

::: {.panel-tabset}

### R

```{r}
#| label: obs-r
cattle_observations <- function(model, data, X, i, t) {
  w <- rep(1, data$n_states)

  y <- data$rams[t, i]
  if (y >= 0) {
    w[1] <- w[1] * ifelse(y == 1, 0, 1)                          # S never tests +
    w[2] <- w[2] * ifelse(y == 1, model$theta_r, 1 - model$theta_r)
  }

  y <- data$faecal[t, i]
  if (y >= 0) {
    w[1] <- w[1] * ifelse(y == 1, 0, 1)
    w[2] <- w[2] * ifelse(y == 1, model$theta_f, 1 - model$theta_f)
  }
  w
}
```

### Julia

```julia
function cattle_observations(model, data, X, i, t)
    w = ones(eltype(model.θʳ), data.n_states)
    for (results, θ) in ((data.rams, model.θʳ), (data.faecal, model.θᶠ))
        y = results[t, i]
        y < 0 && continue                       # not tested
        w[1] *= y == 1 ? 0.0 : 1.0              # S never tests positive
        w[2] *= y == 1 ? θ : (1 - θ)
    end
    w
end
```

:::

::: {.callout-note}
## One thing worth knowing in both languages

`rep(1, data$n_states)` in R, and `ones(eltype(model.θʳ), ...)` in Julia, do the
same job: allocate at the **parameter's** scalar type rather than at `Float64`.

This matters because that type depends on the *blocking*, not the model. When a
test sensitivity is sampled conjugately it is a plain number; when it sits in an
HMC block it is whatever the autodiff backend uses. Writing `zeros(Float64, ...)`
works until someone moves a parameter between blocks, and then throws.

In R the package handles it — `rep(1, n)` is translated appropriately. In Julia
you write it yourself, which is why the idiom is `eltype(model.something)` rather
than a literal type.
:::

## Assembling

::: {.panel-tabset}

### R

```{r}
#| label: data-r
dat <- et_data(
  n_individuals  = n_individuals,
  n_timepoints   = n_timepoints,
  transitions    = trans,
  starting_state = starting_state,
  aggregates     = aggs,
  group          = group,
  observation_process = cattle_observations,
  # your own data, reachable as data$rams / data$faecal above
  extras = list(rams = rams, faecal = faecal))
```

### Julia

```julia
data = epidemic_data(
    n_individuals = n_individuals,
    n_timepoints  = n_timepoints,
    trans_mat     = trans_mat,
    starting_state = starting_state,
    aggregates    = aggs,
    group         = group,
    observation_process = cattle_observations,
    # your own data, reachable as data.rams / data.faecal above
    rams = Rmask, faecal = Fmask)
```

:::

Anything the package does not name is yours, and it never looks inside.

## Parameters and priors

::: {.panel-tabset}

### R

```{r}
#| label: model-r
mod <- et_model(
  data = dat,
  parameters = list(
    alpha   = et_par(et_gamma(1, 1), init = 0.05),
    beta    = et_par(et_gamma(1, 1), init = 0.05),
    m_tilde = et_par(et_gamma(2, 4), init = 4.0),
    nu      = et_par(et_beta(1, 1),  init = 0.10),
    theta_r = et_par(et_beta(1, 1),  init = 0.70),
    theta_f = et_par(et_beta(1, 1),  init = 0.40)),
  # m = m_tilde + 1 keeps 1/m < 1, so the recovery probability can never leave
  # (0, 1) and send the sampler into a DomainError.
  derived = list(m = quote(m_tilde + 1)))
```

### Julia

```julia
@model function cattle_model(data, n_time, n_ind, loglik_fn, obs_loglik_fn)
    α ~ Gamma(1, 1)
    β ~ Gamma(1, 1)
    m̃ ~ Gamma(2, 4)
    m := m̃ + 1.0          # keeps 1/m < 1

    ν  ~ Beta(1, 1)
    θʳ ~ Beta(1, 1)
    θᶠ ~ Beta(1, 1)

    # The whole trajectory is ONE latent block: held constant during the
    # gradients, resampled once per Gibbs sweep by iFFBS.
    X ~ TrajectoryLatent(n_time, n_ind)

    pars = (; α, β, m, ν, θʳ, θᶠ)

    # BOTH halves. Omitting the observation term is a silent modelling bug:
    # every test parameter would then be sampled from its prior.
    @addlogprob! loglik_fn(pars, data, X) + obs_loglik_fn(pars, data, X)
end
```

:::

::: {.callout-important}
## The likelihood has two halves and you need both

`epidemic_loglik` covers the starting state and the transitions.
`epidemic_obs_loglik` covers the observation process. Neither includes the other.

In Julia you write both terms yourself, and forgetting the second is a real bug
that has shipped before: the observation process is used only by the filter, so
every test parameter receives no likelihood information and is effectively
sampled from its prior. The symptom is subtle — the parameters look
prior-shaped — and there is no error.

In R the package emits both whenever an observation process is given, and warns
when one is not.
:::

## The sampler

::: {.panel-tabset}

### R

```{r}
#| label: fit-r
#| eval: false
fit <- et_sample(mod,
  blocks = list(
    et_nuts(c("alpha", "beta", "m_tilde")),
    # The initial infection frequency and the two test sensitivities have
    # closed-form conditionals, so they get conjugate kernels rather than NUTS -
    # an exact independent draw beats a correlated HMC step.
    et_conjugate_initial_state("nu", at = 1, states = c("S", "I")),
    et_conjugate_test_sensitivity("theta_r", y = "rams",   infected_state = "I"),
    et_conjugate_test_sensitivity("theta_f", y = "faecal", infected_state = "I"),
    et_iffbs("X")),
  n_sweeps = 2000, n_burn = 800, n_adapts = 400, seed = 7,
  x_init = X_true)

summary(fit)
```

### Julia

```julia
loglik     = epidemic_loglik(data)
obs_loglik = epidemic_obs_loglik(data)
latent!    = epidemic_latent_sampler(data)

model = cattle_model(data, n_timepoints, n_individuals, loglik, obs_loglik)

spl = Gibbs(
    (:α, :β, :m̃) => NUTS(0.8),
    :ν  => initial_state_kernel(:ν; at = 1, eligible = (X, d, i, t) -> true,
                                states = (1, 2), prior = [1.0, 1.0]),
    :θʳ => test_sensitivity_kernel(:θʳ; Y = Rmask, infected_state = 2),
    :θᶠ => test_sensitivity_kernel(:θᶠ; Y = Fmask, infected_state = 2),
    :X  => iffbs_kernel(latent!; params = (:α, :β, :m, :ν, :θʳ, :θᶠ)),
)

# The aggregates must agree with the STARTING trajectory before the first
# likelihood call. Established once here; the sampler preserves it thereafter.
X0 = copy(X_true)
reset_aggregates!(data)
apply_derived_summaries!((;), data, X0)

chn = AbstractMCMC.sample(StableRNG(7), model, spl, 2000;
                          init = (; X = X0), n_adapts = 400,
                          discard_initial = 800,
                          save_states = (X = :buffer,))
```

:::

In R, `blocks` is optional: omit it and every parameter goes into one NUTS block
with the trajectory sampled by iFFBS. It is spelled out here because the three
conjugate kernels are worth having.

::: {.callout-note collapse="true"}
## Why the trajectory is one block

The hidden trajectory `X` is stored as a **single whole-matrix latent**, not as a
family of indexed variables. That is what lets it be held constant during the
gradient evaluations and resampled once per Gibbs sweep by iFFBS.

It is also the reason this package exists. Written naively inside a model body, a
latent-state sampler re-runs on *every* gradient call — expensive, and because
the sampler is stochastic it makes the gradient itself stochastic, which can
invalidate the inference outright.
:::

## Results

Both versions recover the simulated truth. From the R run:

```
          mean     sd   2.5%    50%  97.5%
alpha   0.0121 0.0032 0.0068 0.0117 0.0197
beta    0.0190 0.0042 0.0117 0.0188 0.0279
nu      0.1324 0.0423 0.0619 0.1286 0.2245
theta_r 0.7328 0.0457 0.6404 0.7348 0.8156
theta_f 0.4656 0.0428 0.3821 0.4653 0.5498
m       6.2078 0.8380 4.7773 6.1505 8.0649
```

| parameter | truth | posterior mean |
|---|---|---|
| `alpha` | 0.01 | 0.0121 |
| `beta` | 0.02 | 0.0190 |
| `m` | 6 | 6.21 |
| `nu` | 0.10 | 0.132 |
| `theta_r` | 0.8 | 0.733 |
| `theta_f` | 0.5 | 0.466 |

All six within three standard errors, from 2000 sweeps in about 49 seconds — on
data where the states are never directly observed.

## Seeing the Julia from R

The generated module is a first-class artefact, not a hidden intermediate. It is
commented, it runs standalone, and it is the on-ramp if you want to move a model
from R to Julia.

```{r}
#| label: source
#| eval: false
et_julia_source(mod)   # print the module this model generates
et_source(fit)         # the module behind a fit
et_eval("...")         # run anything in the session
```

That is how the two halves of this document relate: the R on the left generates
something very close to the Julia on the right.

## Where to go next

* **[The badger model](badger.qmd)** — the same ideas at scale: 2384 individuals,
  four states, six tests, capture-recapture observation, survival, and the
  performance seams this document deliberately ignores.
* `et_check_depends()` — verifies the dependency annotations that let each Gibbs
  block skip likelihood terms it cannot move.
* `et_residuals()` — post-hoc PIT residuals for the mechanisms in your model,
  which is how you find out whether any of this fitted well.
